ExaModels and MadNLP: Open-Source Software Infrastructure for Accelerated Nonlinear Optimization on GPUs

### 2023-08-22 coin-or/main.tex
coin-or letter added

\documentclass{article}

\input{preamble}
\title{ExaModels and MadNLP: Open-Source Software Infrastructure for Accelerated Nonlinear Optimization on GPUs}
\author{Sungho Shin$^\dag$, Francois Pacaud$^\ddag$, and Mihai Anitescu$^\dag$} 
\date{\small
  $^\dag$Mathematics and Computer Science Division, Argonne National Laboratory\\
  $^\ddag$Centre Automatique et Systèmes, Mines Paris - PSL, Paris, France
}
\begin{document}
\maketitle

We are pleased to submit our nominations for the COIN-OR cup, highlighting our contributions of ExaModels and MadNLP in the domain of computational infrastructure for operations research, in the particular area of nonlinear optimization. Our open-source software packages represent a significant advance in addressing the challenges of efficiently solving large-scale nonlinear optimziation problems by harnessing the capabilities of modern GPU hardware. In this document, we highlight the key challenges and opportunities we are facing for the development of optimization software in 2023, especially in the context of the advent of accelerated computing. We also explain how ExaModels and MadNLP jointly address the key technical challanges and harness the benefits that modern accelerated computing has to offer. We also highlight several performance improvement we have achieved in terms of solving large-scale, real-world nonolinear optimziation problems. Finally, we conclude by discussing how our development sets the stage for the next-generation nonlinear optimization solver development and opens up new possibilities in various applications.

\paragraph{Challenges and Significance}
The computational landscape of today presents us with immense challenges and potential opportunities, especially with Graphics Processing Units (GPUs) at our disposal. Notably, NVIDIA GPUs have enabled great stride in the scalabiltiy of deep learning and achieved remarkable success in building large-scale AI models. Also, in the public doamin, most of the computational power for the next generation exascale HPC systems, such as Frontier and Aurora, come from GPU accelerators. While the GPUs have enabled great success in different scientific computing areas, such as machine learning and high-fidelity simulations, utilizing GPUs within nonlinear programming (NLP) algorithms has been hindered by challenges like sparse automatic differentiation (AD) and sparse linear solver routines, which do not seamlessly translate to GPU architecture. The key issue is that the conventional sparse AD and matrix factorization algoirthms are built for CPUs, and they heavily rely on the serial computation.

While GPU computation can trivially accelerate several parts of the optimization process---especially various internal computations within the optimization solver---the sluggish data transfer between host and device memory hampers the ad-hoc implementation of GPU accelerations (Fig. \ref{fig:memory}). To fully leverage the potential offered by modern GPU hardware, it becomes imperative to have a comprehensive computational framework for optimization on GPUs. That is, we need an AD/algebraic modeling framework, sparse linear solvers, and NLP solvers that can operate entirely on the GPU. Specifically, for the best performance, both the problem data and the solver's intermediate computational data must be exclusively resident within the device memory, with the majority of operations executed on the GPU. This necessitates the development of {\it a comprehensive nonlinear optimization solution framework}, which performs all the necessary computations---automatic differentiation, nonlinear optimization, and linear algebra---exclusively on GPUs. ExaModels and MadNLP rise to these challenges, carving out a pathway to fully harness the power of GPUs for operaitons research and nonlinear optimization.

\begin{figure}[t]
  \centering
  \scalebox{.87}{
  \begin{tikzpicture}[remember picture, scale=.9, font=\small]

@@

    \def\rows{10}
    \def\cols{10}
    \def\elementwidth{2}
    \foreach \xshift/\yshift in {.25/1.5, 5.75/1.5} {
      % Draw Grid
      \foreach \i in {0,...,\rows} {
        \draw (\xshift, \yshift+\i*\elementwidth/\rows) -- (\xshift+\elementwidth, \yshift+\i*\elementwidth/\rows);
      }
      \foreach \i in {0,...,\cols} {
        \draw (\xshift+\i*\elementwidth/\cols, \yshift) -- (\xshift+\i*\elementwidth/\cols, \yshift+\elementwidth);
      }
    }
  \end{tikzpicture}
  }
  \includegraphics[width=.48\textwidth]{speedup-sol.pdf}
  \label{fig:simd}
  \caption{A schematic of host (CPU) and device (GPU) memory structure (left) and speedup achieved for AC optimal power flow problems (right).}\label{fig:memory}
\end{figure}

\paragraph{Developed Software Tools}
We have developed a comprehesnive nonlinear optimization framework by implementing our algebraic modeling/automatic differentiation tool ExaModels and nonlinear optimization solver MadNLP, while linear algebra computation is currently performed by the external cuSOLVER library.

{\it ExaModels}: SIMD Abstraction for NLPs: ExaModels pioneers the implementation of a groundbreaking single-instruction, multiple-data (SIMD) abstraction for NLPs. This revolutionary approach unlocks the potential for efficient parallel automatic differentiation on GPUs. The essence of this innovation lies in preserving the parallelizable structure within model equations, transforming derivative evaluations into streamlined operations. By enabling efficient computation of derivatives on GPUs, ExaModels offers an exceptional performance boost for power systems analysis.

The optimization models are most of the times implemented by the researchers and practitioners who do not necessarily understand the operations 
ExaModels provides a new platform for scalable nonlinear optimization. Our numerical results suggest that the conventional algebraic modeling systems are limited in terms of efficiently computing the derivates of the model equations (Table \ref{tab:num}), and it does not allow performing the operations on GPUs (Table \ref{tbl:portabiltiy}). ExaModels provide a framework that allows the user to conveniently inform the AD backend the parallelizable structure. Writing the models in this way will enable the GPU compatibility

{\it MadNLP}: Condensed-Space IPM with Inequality Relaxation: MadNLP introduces an equally transformative approach through its condensed-space interior-point method (IPM) with an inequality relaxation strategy. This strategy deftly overcomes the challenges posed by sparse matrix factorization on GPUs. The relaxation of equality constraints and condensation of the Karush-Kuhn-Tucker (KKT) system establishes positive definiteness, paving the way for the utilization of highly efficient linear solvers. MadNLP thus unveils a remarkable solution to traditionally intractable problems in GPU-based optimization.

While MadNLP has started as a port of Ipopt on Julia Language, now marking as the fourth year since its initial development, MadNLP has become a lot more versatile solver than the previously developed nonlinear optimization solvers. The key features include (1) being able to solve dense optimzation problems, (2) ability to handle different forms of KKT systems, (3) being able to handle diverse array data types (most notably, device arrays), and (4) exploit various Hessian approximation strategies (BFGS variants and limited-memory BFGS methods).

@@

The significance of these contributions is not merely theoretical. Empirical results speak volumes about the impact of ExaModels and MadNLP. When applied to AC OPF problems, the performance enhancements achieved are nothing short of extraordinary. The GPU-based solutions attain speedups of over 20 times compared to CPU-based counterparts, showcasing the profound potential of these software packages in power systems analysis. Their prowess surpasses even established tools interfaced with CPUs, marking a paradigm shift in computational efficiency.

\cite{shin2023accelerating}

\begin{table}[t]
  \begin{center}
    \include{table}
    \vspace{-1em}
    \footnotesize
    $^\dag$Wall time (sec) measured by Julia. $^\ddag$CPU time (sec) reported by Ipopt.  \caption{Numerical Performance of ExaModels and MadNLP}
  \end{center}
  \label{tab:num}
\end{table}

@@

While NVIDIA seems to be most capable in computing power and mature in terms of the implementation of the low-level operations (especially, the CUDA libraries), we aim to support all the other alternative architectures. This is because quite a few next-generaiton exascale high-performance computing architectures will be based on AMD and Intel accelerator devices. Also, as an open-source software, we aim to be the most generic software tool, not the most efficient one. By supporting diverse platforms, we hope to spark the attention of the chip makers in the operations research area and encourage better support for the software infrastructure for implmenting operations research software.

Table \ref{tbl:portabiltiy} summarizes the current state of the portability of our tools and other existing open-source software tools. Notably, ExaModels.jl already support all major architectures, whereas MadNLP.jl only support NVIDIA GPUs. Nonetheless, we have clear path forward in supporting all the major accelerator architectures, including NVIDIA, AMD, and Intel GPUs, while Apple GPUs are somewhat limited due to their inability to perform double precision operations. This is due to the fact that all of our GPU kernels are implemented via KernelAbstractions.jl, a julia package enabling portable implementation of GPU kernels. In this way, the GPU kernels for diverse accelerator architectures can be run based on a sam source code. Using this strategy, recently ExaModels.jl has demonstrated its capabiltiy to run on 4 different accelerator architectures (multi-thread CPUs and NVIDIA, AMD, and Intel GPUs). In the coming years, we plan to exand MadNLP's support for various accelerator architectures.

However, at the same time, it is absolutely crucial to have efficient and reliable, and preferrably also portable sparse linear sovler routines. Currently, we rely on CUSOLVER library for solving the linear systems. While vendor-implmented libraries are expected to have the best performance via high level of optimization, it would be beneficial for the community to have the open-source portable linear solvers. By showcasing the capability of solving high-stake nonlinear optimziation problems on GPUs, we can attract the attention of the numerical linear algebra community so that we can encourage the development of capable open-source sparse linear solvers. 
1
\begin{table}[t]
  \begin{center}
    \scalebox{0.72}{
      \begin{tabular}{|c|c|cccccc|}
        \hline
        &&CPU (single) &CPU (multi)& NVIDIA & AMD & Intel& Apple\\
        \hline
        \multirow{3}{*}{Modeling Platforms}& {\bf AMPL}& \cmark& \xmark& \xmark& \xmark& \xmark& \xmark\\
        &{\bf JuMP}& \xmark& \cmark& \xmark& \xmark& \xmark& \xmark\\
        &{\bf ExaModels}& \cmark& \cmark& \cmark& \cmark& \cmark& \xmark\\ 
        \hline
        \multirow{2}{*}{Solvers} & {\bf Ipopt}& \cmark& \xmark& \xmark& \xmark& \xmark& \xmark\\
        & {\bf MadNLP} & \cmark& \xmark& \cmark& \xmark& \xmark& \xmark\\
        \hline
      \end{tabular}
    }
  \end{center}
  \caption{The current status on the portability of the nonlinear optimization frameworks.}
  \label{tbl:portabiltiy}
\end{table}

\paragraph{Impacts in Various Applications}

{\it Energy Infrastructures.} The resillience We have recently showcased the capability of MadNLP with multiple GPUs for solving extremely-large scale security-constrained AC OPF problems.

{\it 1}

{\it Machine Learning Surrogate Models.}

\paragraph{Closing Remarks}
ExaModels and MadNLP embody the spirit of innovation and progress that the COIN-OR cup seeks to recognize. By addressing critical challenges and introducing transformative strategies, these packages extend the frontiers of computational infrastructure for operations research. We are confident that their exceptional contributions will resonate within the operations research community and beyond, redefining the future of power systems analysis and optimization. It is with great enthusiasm that we nominate ExaModels and MadNLP for the COIN-OR cup, recognizing their exceptional impact on the field.

\bibliography{main}

\end{document}

>>>>

\input{preamble}
\usepackage{wrapfig}
\title{ExaModels and MadNLP: Open-Source Software Infrastructure for Accelerated Nonlinear Optimization on GPUs}
\author{Sungho Shin$^\dag$, Francois Pacaud$^\ddag$, and Mihai Anitescu$^\dag$} 
\date{\small
  $^\dag$Mathematics and Computer Science Division, Argonne National Laboratory\\
  $^\ddag$Centre Automatique et Systèmes, Mines Paris - PSL, Paris, France
}
\begin{document}
\maketitle

We are delighted to submit our nominations for the COIN-OR Cup,
highlighting our contributions of {\bf ExaModels} and {\bf MadNLP} in the field of
computational infrastructure for operations research, specifically in
the area of nonlinear programming (NLP). Our open-source software
packages represent a significant breakthrough in efficiently solving
large-scale NLPs by leveraging the capabilities of modern GPU
hardware.  In this document, we address the primary challenges and
opportunities for optimization software development in 2023,
particularly within the context of accelerated computing. We elucidate
how ExaModels and MadNLP collectively address technical obstacles and
capitalize on the advantages offered by modern accelerated
computing. Furthermore, we emphasize the remarkable performance
enhancements achieved in solving real-world nonlinear optimization
problems at scale. A noteworthy example of such progress is our recent
accomplishment in accelerating alternating current (AC) optimal power
flow (OPF) problems by up to ten times compared to state-of-the-art
tools. Finally, we discuss how our development lays the foundation for
the advancement in various topics in operations research and unlocks
new possibilities.

\paragraph{Challenges and Opportunities}
In recent years, the scalability of deep learning has experienced
significant advancements, primarily due to the utilization of NVIDIA
GPUs, which have demonstrated remarkable success in constructing
large-scale AI models. Additionally, in the public domain, GPU
accelerators have provided a substantial portion of the computational
power for next-generation leadership HPC systems such as Frontier and
Aurora, achieving the milestone of one exaflop per second.  While GPUs
have proven highly effective in various scientific computing areas
like machine learning and high-fidelity simulations, integrating them
into constrained nonlinear programming (NLP) algorithms has encountered
considerable challenges. Specifically, the seamless translation of
sparse automatic differentiation (AD) and sparse linear solver
routines to GPU architecture has been recognized to be particularly
challenging.  This is because conventional sparse AD and matrix
factorization algorithms are designed for CPUs and heavily rely on
serial computation, making them unsuitable for the single-instruction,
multiple-data (SIMD) parallelism employed by GPU architectures (see
Figure \ref{fig:simd}, left).

While GPU computation can accelerate several parts of the optimization
process straightforwardly, the slow data transfer between host and
device memory hinders the ad-hoc implementation of GPU accelerations
(Figure \ref{fig:simd}, right). To fully exploit the potential offered
by modern GPU hardware, it is crucial to develop a comprehensive
computational framework for optimization on GPUs. This requires
keeping the problem data and the solver's intermediate computational
data reside exclusively within device memory, with the majority of
operations executed on the GPU. Consequently, the development of a
comprehensive NLP solution framework that performs
all necessary computations, including AD, optimization, and linear
algebra, exclusively on GPUs becomes imperative.  To address these
challenges, ExaModels and MadNLP offer a comprehensive solution
framework for NLP. They enable AD and NLP solution
exclusively on GPUs, while linear algebra computation is performed by
the external library (e.g., cuSOLVER), allowing for the full
utilization of GPUs.

\begin{figure}[t]
  \centering 
  \scalebox{.82}{
    \begin{tikzpicture}[remember picture, scale=.9, font=\small]
    \node[draw,align=left] (I) at (0,2.75) {
      $\bullet$ Instruction $g(\cdot,\cdot)$\\
      $\bullet$ Pointer to array $x$
    };
    \node[draw] at (-4,0) {
      \tikz{
        \foreach\x in {1,2,3,4,5,7}{
          \node (A\x) at (0,-\x/2) {$s_{\x}$};
        }
        \node at (0,-2.9) {$\vdots$};
      }
    };
    \node[draw] at (0,0) {
      \tikz{
        \foreach\x in {1,2,3,4,5,7}{
          \node (B\x) at (0,-\x/2) {$\text{PU}_{\x}$};
        }
        \node at (0,-2.9) {$\vdots$};
      }
    };
    \node[draw] at (4,0) {
      \tikz{
        \foreach\x in {1,2,3,4,5,7}{
          \node (C\x) at (0,-\x/2) {$y_{\x}$};
        }
        \node at (0,-2.9) {$\vdots$};
      }
    };
    \foreach\x in {1,2,3,4,5,7}{
      \draw[->] (A\x.east) -- (B\x.west);
      \draw[->] (B\x.east) -- (C\x.west);
      \draw[->] (I.west) to [out=-135,in=180] (B\x.west);
    }
    \node at (-4,-2.5) {Input Data};
    \node[align=center] at (0,-2.5) {Processing Units\\(CPUs or GPUs)};
    \node at (4,-2.5) {Output Data};
  \end{tikzpicture} 
  }
  \scalebox{.87}{
  \begin{tikzpicture}[remember picture, scale=.9, font=\small]

@@

    \def\rows{10}
    \def\cols{10}
    \def\elementwidth{2}
    \foreach \xshift/\yshift in {.25/1.5, 5.75/1.5} {
      % Draw Grid
      \foreach \i in {0,...,\rows} {
        \draw (\xshift, \yshift+\i*\elementwidth/\rows) -- (\xshift+\elementwidth, \yshift+\i*\elementwidth/\rows);
      }
      \foreach \i in {0,...,\cols} {
        \draw (\xshift+\i*\elementwidth/\cols, \yshift) -- (\xshift+\i*\elementwidth/\cols, \yshift+\elementwidth);
      }
    }
  \end{tikzpicture}
  }
  % \includegraphics[width=.48\textwidth]{speedup-sol.pdf} 
  \label{fig:simd}
  \caption{Schematics of host and device memory architectures (left) and SIMD parallelism (right).}\label{fig:memory}
\end{figure}

\paragraph{ExaModels}

ExaModels is an algebraic modeling and AD system embedded in the Julia
Language, specialized for the SIMD abstraction of nonlinear
programs. This system is designed based on the repetitive structure
commonly found in large-scale NLPs. For example, AC OPF problems,
which may have millions of variables and constraints, can be expressed
using just 15 computational patterns. By enforcing the creation of
models through specifying repetitive patterns, ExaModels preserves the
parallelizable structure within model equations and facilitates
parallel automatic differentiation. Writing models in ExaModels
automatically enables GPU compatibility, which is especially
beneficial as optimization models are often implemented by domain
experts who may not fully understand the internals of algebraic
modeling systems and NLP solvers.  This makes ExaModels a powerful
modeling platform for scalable computations.

By enabling efficient computation of derivatives on GPUs, ExaModels
provides exceptional performance, surpassing existing tools like AMPL
(AD backend of Pyomo) and JuMP by more than two orders of
magnitude. Our numerical results demonstrate that conventional
algebraic modeling systems struggle to efficiently compute derivatives
of model equations (Table \ref{tab:num}), and they are incompatible
with GPU operations (Table \ref{tbl:portability}). In contrast,
ExaModels provides a convenient framework for users to inform the AD
backend about the parallelizable structure, and in turn, efficient,
parallelized AD can be performed in the backend.

\paragraph{MadNLP}

MadNLP is a versatile NLP solver capable of
handling various problem data structures. By leveraging the modularity
provided by the key features of the Julia Language, such as multiple
dispatch and just-in-time compilation, MadNLP offers great flexibility
in handling different types of data structures while retaining the
sophistication of the algorithm. Importantly, with this approach, we can avoid
re-implementing interior point solver every time for each data types,
which in turn ensures that the operations performed are mathematically
equivalent to those in the mature, extensively tested existing code
base. MadNLP has started as a port of Ipopt (an open-source nonlinear
optimziation solver running on CPUs) in the Julia Language, but over
the past four years, it has expanded beyond its initial scope and now
offers a range of capabilities not supported by state-of-the-art NLP
solvers. These include solving dense optimization problems, handling
different numerical precisions, working with different forms of KKT
systems, managing diverse array data types (including GPU device
arrays), and utilizing various Hessian approximation strategies, such
as different variants of BFGS methods.

Recently, we have discovered that the condensed-space interior point
method (IPM) with an inequality relaxation strategy is particularly
effective for solving large-scale NLPs, like AC optimal power flow, up
to moderate precisions. This strategy efficiently addresses the
challenges posed by sparse matrix factorization on GPUs, as the
condensed KKT systems can be factorized without numerical pivoting,
which has previously hindered GPU utilization in the large-scale
optimization regime.  Furthermore, MadNLP offers diverse approaches
for handling KKT systems with special structures, via reduction
strategies (citation needed) and Schur complement decomposition
strategies (citation needed).

@@

Our numerical results, summarized in Table \ref{tab:num} (for detailed
results, please refer to \cite{shin2023accelerating}), clearly
demonstrate the significant potential of using GPUs. Specifically,
when applied to AC OPF problems, which are one of the most important
applications of NLP, we have achieved significant performance
enhancements. The GPU-based solutions achieve speedups of over 10
times compared to CPU-based state-of-the-art tools. Notably, ExaModels
enables derivative evaluations that are up to two orders of magnitude
faster, and the use of the condensed-space IPM strategy in MadNLP
greatly improves linear algebra computation speed. Furthermore, we
have also explored the utilization of these methods in reduced-space
OPF \cite{} and model predictive control (MPC) \cite{}, where we have
achieved similar degrees of speedup.

\begin{table}[t]
  \begin{center}
    \include{table}
    \vspace{-1em}
    \footnotesize
    $^\dag$Wall time (sec) measured by Julia. $^\ddag$CPU time (sec) reported by Ipopt.  \caption{Numerical Performance of ExaModels and MadNLP for solving AC OPF problems}
  \end{center}
  \label{tab:num}
\end{table}

@@

While NVIDIA currently leads in terms of computing power and mature
implementation of low-level operations, especially with CUDA
libraries, we aim to provide support for alternative architectures as
well. This is crucial because many next-generation exascale
high-performance computing architectures will be based on AMD and
Intel accelerator devices. Furthermore, by supporting diverse
platforms, we hope to drive the attention of chip makers towards the
operations research field and encourage better support for software
infrastructure in implementing operations research software.

Table \ref{tbl:portability} provides an overview of the current state
of portability for our tools and other existing open-source software
tools. ExaModels.jl already supports all major architectures, while
MadNLP.jl currently only supports NVIDIA GPUs. However, we have a
clear roadmap to extend support to all major accelerator
architectures, including NVIDIA, AMD, and Intel GPUs. The portability
is enabled by KernelAbstractions.jl, a Julia package that facilitates
the portable implementation of GPU kernels.

\begin{table}[t]
  \begin{center}
    \scalebox{0.72}{
      \begin{tabular}{|c|c|cccccc|}
        \hline
        &&CPU (single) &CPU (multi)& NVIDIA & AMD & Intel& Apple\\
        \hline
        \multirow{3}{*}{Algebraic Modeling Platforms}& {\bf AMPL}& \cmark& \xmark& \xmark& \xmark& \xmark& \xmark\\
        &{\bf JuMP}& \cmark& \xmark& \xmark& \xmark& \xmark& \xmark\\
        &{\bf ExaModels}& \cmark& \cmark& \cmark& \cmark& \cmark& \xmark\\ 
        \hline
        \multirow{2}{*}{NLP Solvers} & {\bf Ipopt}& \cmark& \xmark& \xmark& \xmark& \xmark& \xmark\\
        & {\bf MadNLP} & \cmark& \xmark& \cmark& \xmark& \xmark& \xmark\\
        \hline
      \end{tabular}
    }
  \end{center}
  \caption{The current status on the portability of the NLP frameworks.}
  \label{tbl:portabiltiy}
\end{table}

\paragraph{Expected Impacts in Applications}

\begin{itemize}
\item \textit{Energy Infrastructures}: Enhancing the scalability of
optimization methods is crucial for improving the resilience of energy
system operations. We have demonstrated the ability of MadNLP with
multiple GPUs to solve large-scale security-constrained AC OPF
problems effectively. This showcases the potential of ExaModels and
MadNLP to enhance the scalability of NLP by utilizing GPU
capabilities.

\item \textit{Model Predictive Control}: MPC has been widely used for
optimization-based control in various settings, such as chemical
processes, energy systems, and robotics. However, the effectiveness of
MPC is often limited by the efficiency of optimization solvers,
especially when short sampling times are required. Our results have
shown that MadNLP running on GPUs can solve such problems with high
efficiency \cite{}. Moving forward, we plan to implement the extension
packages for control problems so that control community can take
advantage of accelerated NLP frameworks.

\item \textit{Machine Learning Surrogate Models}: An exciting future
direction is the integration of neural network surrogate models into
constrained optimization algorithms. GPU acceleration plays a crucial
role in these applications, considering the complexity of real-world
AI models. MadNLP offers the necessary capabilities through the native
handling of device arrays as well as dense KKT systems with various
quasi-Newton strategies. We intend to explore these problems further
and provide extensions for surrogate modeling with GPU acceleration.

\end{itemize}

\paragraph{Closing Remarks} ExaModels and MadNLP provide computational
infrastructures for operations research that align with the era of
accelerated computing, which we believe the COIN-OR cup aims to
recognize. By addressing critical challenges and introducing new
strategies, these packages significantly extend the boundaries of
computational infrastructure for operations research. We are confident
that these contributions will resonate within the operations research
community and beyond, opening up new possibilities in applications
such as energy systems, control, and machine learning. With great
enthusiasm, we nominate ExaModels and MadNLP for the COIN-OR cup, in
recognition of their impact on the field.

\bibliography{main}
\end{document}
